Nanophotonics with optical phonons in two dimensions

### 2018-11-09 paper.tex
discussion on anomalies in monolayer vs bulk

where $m,n$ are states of the phononic Fock space of the monolayer, $P_{i,mn} = \sum\limits_{\kappa,j}\langle m | u_{\kappa,j} | n \rangle$ are matrix elements of the polarization associated with phonon modes, $E_{m (n)}$ is the energy of state $m$ $(n)$, and $\beta$ is the inverse temperature. Evaluating the contribution of optical phonons in the harmonic approximation to the polarization-polarization response, one finds that the response function is given by: 

@@

We now re-write $\epsilon_{\mathrm{RPA}}$ in a way that is expressed purely in terms of the optical phonon dispersion, derive the conductivity of the monolayer, and derive a universal form for the dispersion of 2D LO phonons in the long-wavelength limit. From Equation (8), we may write Equation (7) as
\begin{equation}
\epsilon_{\mathrm{RPA}}(\omega) = \epsilon_{\mathrm{env}}\left( 1 + \frac{\omega^2_{\mathbf{q}\mathrm{LO}}-\omega^2_{\mathbf{q}\mathrm{TO}}}{\omega^2_{\mathbf{q}\mathrm{TO}}-\omega^2}\right).
\end{equation}
Similarly, the conductivity is prescribed by $\sigma(\mathbf{q},\omega) = -\iu \omega \Pi(\mathbf{q},\omega)$. We are now in a position to derive a universal form for the RPA dielectric function of a 2D polar slab in terms of three phenomenological parameters. In many cases [ref], the Born charges are independent of $q$ for small $q$, and thus the dispersion can be written as
\begin{equation}
\omega_{\mathbf{q},\mathrm{LO}} = \sqrt{\omega^2_{\Gamma \mathrm{TO}}+\mathcal{S}q} \approx \omega_{\Gamma \mathrm{TO}} + v_g q,
\end{equation}
where $\mathcal{S}$ is a microscopic constant determined by the Born charges of the monolayer, $v_g$ is the group velocity of the LO mode, and $\omega_{\Gamma \mathrm{TO}}$ is the TO phonon frequency at the $\Gamma$ point. The group velocity is specified in terms of microscopic parameters via 
\begin{equation}
v_g = \frac{e^2 }{4\epsilon_0 \epsilon_{\mathrm{env}}\omega_{\Gamma TO}\Omega}\Big|\sum\limits_{\kappa,j}Z^r_{\kappa,ij}\eta_{\kappa,j}  \Big|^2. 
\end{equation}
Thus, in the long wavelength limit, we have the following universal parameterization of the RPA dielectric function of a polar monolayer:
\begin{equation}
\epsilon_{\mathrm{RPA}}(\omega) = \epsilon_{\mathrm{env}}\left( 1 + \frac{2v_g\omega_{\mathbf{q}\mathrm{TO}}}{\omega^2_{\mathbf{q}\mathrm{TO}}-\omega^2+\iu \omega \tau^{-1}}\right),
\end{equation}
where we have also phenomenologically included the phonon dissipation rate $\tau^{-1}$, consistently with  a relaxation-time prescription. Thus, all of the optical properties associated with LO phonons is captured by the TO phonon frequency, the slope of the LO dispersion at $\Gamma$, and the dissipation $\tau^{-1}$.

@@

@@

We now move to analyze key figures of merit of 2D LO phonons.

>>>>

where $m,n$ are states of the phononic Fock space of the monolayer, $P_{i,mn} = \sum\limits_{\kappa,j}Z_{\kappa,ij}\langle m | u_{\kappa,j} | n \rangle$ are matrix elements of the polarization associated with phonon modes, $E_{m (n)}$ is the energy of state $m$ $(n)$, and $\beta$ is the inverse temperature. Evaluating the contribution of optical phonons in the harmonic approximation to the polarization-polarization response, one finds that the response function is given by: 

@@

We now re-write $\epsilon_{\mathrm{RPA}}$ in a way that is expressed purely in terms of the optical phonon dispersion, derive the conductivity of the monolayer, and derive a universal form for the dispersion of 2D LO phonons in the long-wavelength limit. From Equation (8), we may write Equation (7) as
\begin{equation}
\epsilon_{\mathrm{RPA}}(\omega) = \epsilon_{\mathrm{env}}\left( 1 + \frac{\omega^2_{\mathbf{q}\mathrm{LO}}-\omega^2_{\mathbf{q}\mathrm{TO}}}{\omega^2_{\mathbf{q}\mathrm{TO}}-\omega^2}\right).
\end{equation}
Similarly, the conductivity is prescribed by $\sigma(\mathbf{q},\omega) = -\iu \omega \Pi(\mathbf{q},\omega)$. We are now in a position to derive a universal form for the RPA dielectric function of a 2D polar slab in terms of three phenomenological parameters. In many cases [ref], the Born charges are independent of $q$ for small $q$, and thus the dispersion can be written as
\begin{equation}
\omega_{\mathbf{q},\mathrm{LO}} = \sqrt{\omega^2_{\Gamma \mathrm{TO}}+\mathcal{S}q} \approx \omega_{\Gamma \mathrm{TO}} + v_g q,
\end{equation}
where $\mathcal{S}$ is a microscopic constant determined by the Born charges of the monolayer, $v_g$ is the group velocity of the LO mode, and $\omega_{\Gamma \mathrm{TO}}$ is the TO phonon frequency at the $\Gamma$ point. The group velocity is specified in terms of microscopic parameters via 
\begin{equation}
v_g = \frac{e^2 }{4\epsilon_0 \epsilon_{\mathrm{env}}\omega_{\Gamma TO}\Omega}\Big|\sum\limits_{\kappa,j}Z^r_{\kappa,ij}\eta_{\kappa,j}  \Big|^2. 
\end{equation}
Thus, in the long wavelength limit, we have the following universal parameterization of the RPA dielectric function of a polar monolayer:
\begin{equation}
\epsilon_{\mathrm{RPA}}(\omega) = \epsilon_{\mathrm{env}}\left( 1 + \frac{2v_g\omega_{\mathbf{q}\mathrm{TO}}}{\omega^2_{\mathbf{q}\mathrm{TO}}-\omega^2+\iu \omega \tau^{-1}}\right),
\end{equation}
where we have also phenomenologically included the phonon dissipation rate $\tau^{-1}$, consistently with  a relaxation-time prescription. Thus, all of the optical properties associated with LO phonons is captured by the TO phonon frequency, the slope of the LO dispersion at $\Gamma$, and the dissipation $\tau^{-1}$. In Table 1, we tabulate these parameters for a few polar monolayers whose phonon band-structures have been calculated in the literature: hBN, MoS$_2$, and ...

@@

\section{Properties of 2D LO phonon modes in monolayer hBN}

In Fig. 2a, we present the dispersion relation, operationally defined through the poles of the Fresnel reflection coefficient for $p$-polarized waves, which for a monolayer of conductivity $\sigma(\mathbf{q},\omega) = -\frac{2\iu \omega}{\epsilon q}\left(\epsilon_{\mathrm{RPA}} - \epsilon_{\mathrm{env}} \right)$ below vacuum and on top of a substrate of permittivity $\epsilon_{\mathrm{s}}$, is given by
\begin{equation}
r_p(\mathbf{q},\omega) = \frac{(\epsilon_{\mathrm{s}}-1)-\frac{\sigma(\mathbf{q},\omega) q}{\omega\epsilon_0}}{(\epsilon_{\mathrm{s}}+1)-\frac{\sigma(\mathbf{q},\omega) q}{\omega \epsilon_0}}.
\end{equation}
In Figs. 2b, we show the confinement factor (or effective mode index) of the 2D LO phonon mode, defined by $n_{\mathrm{eff}} = \frac{qc}{\omega}$. The confinement factor is a key figure of merit in many nanophotonic applications involving quantum emitters, because as was shown in [Rivera], all interactions involving emission of electromagnetic energy by emitters, such as electric and magnetic dipole emission, multipole emission, phosphorescence, and multiphoton emission, scale as some power of this confinement factor. Interestingly, in the case of a monolayer, the confinement grows very rapidly in frequency, due to the extremely low group velocity of 2D LO phonons, which is a remarkable four orders of magnitude slower than the speed of light.  In particular, at frequencies of 1450 $cm^{-1}$, the LO phonon already has a confinement factor in excess of conf, corresponding to a phonon wavelength of 6896.55/conf nm, which is much shorter wavelength than any phonon polariton thus measured so far, as well as shorter wavelength than any plasmon that has been measured in the mid-IR, such as in graphene. In fact, this short wavelength is mostly out of the range of measurability in scattering near-field microscope (SNOM) measurements, which with current tip radii, are limited to measuring wavelengths in excess of 60 nm. This 2D LO phonon can be measured with SNOM at frequencies close to the TO phonon frequency, where the confinement is less, but this comes at the price of very high dissipation and correspondingly low propagation quality factor, as illustrated in Fig. 2c. This is simply because frequencies near the TO phonon, which is strongly absorbing, correspond to the highest dissipation (measured by $\text{Re }\sigma$ or $\text{Im }\epsilon_{\mathrm{RPA}}$). These considerations imply that access to the lower-loss and higher-confined portions of the phononic waves, in the absence of a sharp tip, requires either a free electron probe, as in EELS, or a quantum emitter probe, such as a molecule or quantum well [cite people].

In what remains of this section, we discuss briefly ``anomalies", by which we mean features of the 2D LO phonon dispersion that cannot be predicted by considering the electromagnetic modes of a very thin slab of material whose dielectric function is that of the bulk polar material (with finite LO-TO splitting). For example, the long-wavelength linear dispersion of the 2D LO phonon mode given by Equation (10) is not an anomaly, as this would be predicted by taking a sub-nm thick slab of a polar dielectric material. However, the precise slope cannot be predicted, as it would depend on the thickness used for the slab, which is ill-defined, as one will find that it does not correspond to the thickness of the boron or nitrogen atoms. One immediate anomaly is that in the monolayer, there is only one LO phonon, and thus only one branch of TM modes. This is contrary to a thin slab of hexagonal boron nitride, which is hyperbolic $\epsilon_{zz} > 0, \epsilon_{xx} = \epsilon_{yy} < 0$, and thus has many branches of TM modes spaced in wavevector by an amount proportional to $d^{-1}$, with $d$ the film thickness. Such additional TM modes would be incorrectly predicted by extrapolating the bulk to a film. Another anomaly comes from the fact that the dispersion relation of the 2D LO phonon is non-monotonic, and the frequency starts to decrease for sufficiently large wavevector. This can never happen by extrapolating to a monolayer a dielectric function which has finite LO-TO splitting.

@@

We now move to analyze t
